The hydra and the rooster

### 2020-07-28 latex/chapter-ordinals.tex
redaction introduction to ordinals

@@

It would be interesting to specify at the most generic level, what is a zero, a successor or a limit ordinal.

%%%% ICI ICI %%%%%%%%%%%

More complex examples will be developped in further sections or chapters. For instance,
a notation system for ordinals less than $\epsilon_0$ is presented in chapter~\ref{chap:T1}.

@@

It is sometimes useful to compare two ordinal notations with respect to expressive power
(the segment of ordinals one they represent).

The following class specifies a strict inclusion of segments. The notation \texttt{OA} describes a segment $[0,alpha($, and \texttt{OB} is a larger segment (which contains a notation for $\alpha$). We require that the comparison functions of the two notation systems are compatible.

@@

\begin{todo}
pointers to examples 
\end{todo}

Finally, it may be interesting to compare an ordinal notation with the more theoretical model from Schütte (well, at least with our formalization of thhat model). This would be a relative proof of correctenss of our ordinal  notations.

@@

For instance, the following theorem tells that \texttt{Epsilon0}, our otation system for the segment $[0,\epsilon0)$ is correct with respect to the countqable ordinal $\epsilon_0$
(see chapter~\ref{chap:schutte} for more details).

@@

% ICI

\begin{todo}
Present instances of Ordinal notation.  
\end{todo}

It is usual to classify ordinal numbers in three pairwise disjoint categories: 
\begin{itemize}
\item The category containing only the least ordinal (called $0$),
\item The \emph{successors}: $y$ is a successor of $x$ if $x<y$ and there is no $z$ such that $x<z<y$. 
\item \emph{Limit}: $x\not=0$ is a limit if $x$ is the least upper bound of the set $\{y :\mathbb{O} | y < x\}$.
\end{itemize}

These notions are defined (for any strict order) in the module 
\href{../src/html/hydras.Prelude.MoreOrders.html}{Prelude.MoreOrders}.

\begin{Coqsrc}
Variable lt: relation A.

 Definition Least  {sto : StrictOrder lt} (x : A):= forall y,  clos_refl A lt x y.

\begin{exercise}
Prove in \coq{} that the three predicates above are pairwise exclusive.  
\end{exercise}

\begin{todo}
Present the method ZeroLimitSucc 
\end{todo}

\subsection{Preliminary Definitions}

The type class \texttt{StrictOrder} is already defined in Standard Library
(module.

\begin{itemize}
\item Let  $\alpha$ be an ordinal; we say that  $\alpha$ is a \emph{successor} if there exists some ordinal  $\beta$ such that 
$\alpha$ is is the least ordinal strictly greater than  $\beta$.

\item We say that an ordinal $\lambda$ is a \emph{limit ordinal} is $\lambda$  is the least upper bound of a stricly increasing sequence of ordinals.
The meta-variable $\lambda$ will be used for denoting  limit ordinals.

\item \index{Maths!Transfinite induction}
An ordinal is either $0$, a limit ordinal or a successor ordinal. This case analysis, as well as \emph{transfinite} ({i.e.}, well-founded) induction is used in many proofs about ordinal numbers.

\item Let $\alpha$ be any ordinal, the \emph{segment} $\mathbb{O}_\alpha$ is the
interval $[0,\alpha)\,=\,\{\beta|0\leq\beta<\alpha\}$.

\item The set  of finite ordinals is a segment isomorphic to the set of natural numbers.
 The first infinite ordinal is the limit ordinal $\omega$.

\item The operations $+$, $\times$ and exponentiation on $\mathbb{N}$ are extended on ordinals numbers. Note that these  extensions are not commutative any more.  For instance $\omega = 1 + \omega \not= \omega + 1$
and $\omega = 2 \times \omega \not= \omega \times 2$.

\item The ordinal $\epsilon_0$ is the least solution of the equation
 \(\alpha=\omega^{\alpha}\).
\end{itemize}

Please note that the set of countable ordinals is not countable.

Without ambiguity we will often denote the segment $\mathbb{O}_\alpha$ by just $\alpha$.

\section{Notation systems for some small ordinals}

The following enumeration  shows  popular notation systems for several ordinals (shown in increasing order with respect to inclusion), and their possible representation in \coq{}.
Please keep in mind that a notation system for a given ordinal $\alpha$ represents 
ordinals \emph{strictly below} $\alpha$.

\subsection{Finite ordinals}
Let $n$ be some natural number. The segment associated with $n$ is the interval 
$)0,n[\,=\,\{0,1,\dots,n-1\}$.

@@

For more information about this pattern, please lokk at the numerous mailing lists and 
FAQs on \coq{}).

@@

\subsection{The first infinite ordinal : \texorpdfstring{$\omega$}{omega}}

Mathematically speaking, the ordinal $\omega$ is (set-theoretically) considered as the well-ordered set of
natural numbers. 
Every beginner in \coq{} must have seen examples using the type \texttt{nat} of natural numbers. \coq's standard 
library already contains a lot of functions and lemmas on Peano arithmetic.

In order to build a notation system for $\omega$, we have just to look for lemmas about strict orders and well-foundedness in type \texttt{nat}.

\begin{Coqsrc}
Require Import Arith Compare_dec Lia Simple_LexProd Ordinal_generic.
Import Relations.

Search (@StrictOrder nat).
Search @Nat.compare.
Search (@well_founded nat).
\end{Coqsrc}

\begin{Coqsrc}
Instance Omega : OrdinalNotation Nat.lt_strorder Nat.compare.
Proof.
 split.
 - apply Nat.compare_spec.
 - apply lt_wf.
Qed.
\end{Coqsrc}

We can also prove that any finite ordinal $i$ can be considered as a sub-segment of $\omega$.

\begin{Coqsrc}
Global Instance FinOrd_Omega (i:nat) :
  SubSegment (FinOrd i) Omega i 
             (fun alpha =>  proj1_sig alpha).
\end{Coqsrc}

\subsubsection{Finite ordinals as members of \texorpdfstring{$\omega+\omega$}{omega+omega}}

It is easy do define a coercion from type \texttt{nat} into our type \texttt{t}.

\begin{Coqsrc}
Definition fin (n:nat) : t := inl n.
Coercion fin : nat >-> t.
\end{Coqsrc}

The range of \texttt{fin} is called the set of \emph{finite} ordinals.

\subsubsection{Definition of \texorpdfstring{$\omega$}{omega}}

We define $\omega$ as the least ordinal which is larger than any finite ordinal.

\begin{Coqsrc}
 Notation  "'omega'"  := (inr 0) : oo_scope. 
\end{Coqsrc}

\begin{exercise}
Show that, for any $\alpha:t$, $\alpha$ is finite iff $\alpha<\omega$.
\end{exercise}

\subsubsection{Building an instance of OrdinalNotation}

The class \texttt{OrdinalNotation} is parameterized by the strictorder $<$ and a correct comparison function.

First, we prove that \texttt{lt} is an instance of \texttt{StrictOrder}.

\begin{Coqsrc}
Instance lt_strorder : StrictOrder lt.
Qed.
\end{Coqsrc}

Then, we define a comparison function by pattern matching over the sum type $t$.

\begin{Coqsrc}
Definition compare (alpha beta: t) : comparison :=
   match alpha, beta with
     inl _, inr _ => Lt
   | inl n, inl p | inr n, inr p => Nat.compare n p
   | inr _, inl _ => Gt
  end.
\end{Coqsrc}

\begin{Coqsrc}
Compute compare 7 omega.
\end{Coqsrc}

\begin{Coqanswer}
= Lt : comparison
\end{Coqanswer}

\begin{Coqsrc}
Compute compare  (inr 2) 42.
\end{Coqsrc}

\begin{Coqanswer}
= Gt : comparison
\end{Coqanswer}

\begin{Coqsrc}
Compute compare (inr 0) omega.  
\end{Coqsrc}

 \begin{Coqanswer}
= Eq : comparison
\end{Coqanswer}

Correctness of the function \texttt{compare} is asserted by the following lemma:

\begin{Coqsrc}
Lemma compare_correct alpha beta :
    CompareSpec (alpha = beta) (lt alpha beta) (lt beta alpha)
              (compare alpha beta).
\end{Coqsrc}

\begin{todo}
  Well foundedness; instance of OrdinalNotation.
\end{todo}

%%%% ICI %%%%

\subsubsection{Limit and successor ordinals}

The notions of limit and successor are defined in the module 
\href{../src/html/hydras.Prelude.Limits_and_Successors.html}{Prelude.Limits\_and\_Successors} for any instance of
\texttt{StrictOrder}.

\index{Maths!Limit ordinals}

Let $<$ be a strict order on a type $A$ .
We say $x:A$ is an  $\omega$-\emph{limit} of  a sequence $s(i), i\in\mathbb{N}$, if
 if $x$ is [strictly] larger than all the $s(i)$, and for any
$y<x$, $y$ is dominated by at least an element of  the sequence $s$.

\noindent\emph{From Module~\href{../src/html/hydras.Prelude.Limits_and_Successors.html}{Prelude.Limits\_and\_Successors}}

\begin{Coqsrc}
Definition  Omega_limit
            {A:Type}{lt : relation A}
           {sto : StrictOrder lt} (s: nat -> A) (x:A)  :=
  (forall i: nat, lt (s i) x) /\
  (forall y, lt y  x -> exists i:nat, lt y (s i)).

Definition Limit   {A:Type}{lt : relation A}
           {sto : StrictOrder lt}  (x:A)  :=
  exists s : nat -> A, Omega_limit s x.
 \end{Coqsrc}

Let $x$ and $y$ be two terms of type $A$, we say that \emph{$y$ is a successor of $x$}
if $x<y$ and there is no $z$ strictly greater than $x$ and  strictly less than $y$.

\begin{Coqsrc}
Definition Successor `{lt : relation A}
           {sto : StrictOrder lt} (y x : A):=
  lt x y /\ (forall z,  lt x z -> lt z y -> False).
\end{Coqsrc}

@@

\begin{exercise}
Prove that $\omega$ is the unique   limit ordinal within the notation system for $\omega+\omega$
\emph{(in this precise notation system! In larger segments of 
ordinals, there will be \emph{a lot} of limit ordinals}).
\end{exercise}

\subsection{Computing in \texorpdfstring{$\omega+\omega$}{omega+omega}}

The simple data type we use for representing the ordinals less than $\omega+\omega$ allows us to compute successors, and to decide properties like \texttt{Successor} or \texttt{Limit}.

\subsubsection{Computing successors}

The following function computes the successor of any ordinal less than $\omega+\omega$, by a simple pattern matching.

\begin{Coqsrc}
Definition succ (alpha : t) := match alpha with
                                 inl i => inl (S i)
                               | inr i => inr (S i)
                               end.

Compute succ omega.
\end{Coqsrc}

\begin{Coqanswer}
  = inr 1
     : nat + nat
\end{Coqanswer}

\begin{Coqsrc}
Lemma lt_succ alpha : alpha < succ alpha.
Proof.
 destruct alpha; constructor; auto with arith. 
Qed.
\end{Coqsrc}

The correctness  of the \texttt{succ} function is asserted by the following lemma.

\begin{Coqsrc}
Lemma Successor_correct alpha beta : Successor beta alpha <-> beta = succ alpha.
\end{Coqsrc}

Being a successor is decidable, thanks to the following boolean function.

\subsubsection{Deciding whether a given ordinal is a limit}

In the same way as successors, one can decide by pattern matching wether a given ordinal (dont' forget less than $\omega+\omega$ !) is a limit.

\begin{Coqsrc}
Definition limitb (alpha : t) := match alpha with
                                     (inr 0) => true
                                   | _ => false
                                   end.

Lemma limitb_correct alpha  : limitb alpha <-> 
                              exists s: nat -> t, Omega_limit s alpha.
\end{Coqsrc}

\begin{todo}
trichotomy zero/succ/limit
boolean functions limitb, succb

\end{todo}
\begin{itemize}
\item $\omega^n$ (for some  integer  $n\geq 2$) : the type of $n$-uples of natural numbers, with the lexicographic product of $n$ copies of \texttt{Nat.lt}.
\item  $\omega^\omega$: the set of nonincreasing sequences of natural numbers, lexicographically ordered (also the set of finite multisets of natural numbers).
\item $\epsilon_0$: The set of terms in Cantor normal form (see Chap.~\ref{chap:T1}).
\item $\Gamma_0$: The set of terms in Veblen normal form.
\end{itemize}

\begin{todo}
The following section has moved from chaper on hydras.
Adapt the redaction to the $\omega^2$ ordinl segment.
\end{todo}

>>>>

\begin{Coqsrc}
Set Printing All.
\end{Coqsrc}

\begin{Coqanswer}
1 subgoal (ID 337)

  ============================
  @on_lt nat Peano.lt Nat.compare Omega (S (S (S (S (S (S O))))))
    (S (S (S (S (S (S (S (S O))))))))
\end{Coqanswer}

@@

It would be interesting to specify at the most generic level, what is a zero, a successor or a limit ordinal. Let $<$ be a strict order on a type $A$.

\begin{itemize}
\item A \emph{least} element is a minorant (in the large sense) of the full set  on $A$,
\item $y$ is a \emph{successor} of $x$ if $x<y$ and there is no element between $x$ and $y$,
\item $x$ is a \emph{limit} if $x$ is not a least element, and for any $y$ such that $y<x$,
 there exists some $z$ such that $y<z<x$.
\end{itemize}

The following definitions are in Library\href{../src/html/hydras.Prelude.MoreOrders.html}{Prelude.MoreOrders}.

\begin{Coqsrc}
Section A_given.
  Variables (A : Type)  (lt: relation A).

Definition Least {sto : StrictOrder lt} (x : A):=
  forall y,  x <= y.

Using these definitions , we can prove the following lemmas:

\noindent\emph{From Module~\href{../src/html/hydras.Prelude.ON_Omega_plus_omega.html}{Prelude.ON\_Omega\_plus\_omega}}

\begin{Coqsrc}
 Lemma omega_is_limit : Limit omega.

Lemma limit_is_omega alpha : Limit alpha -> alpha = omega.

Regarding successors, let us introduce the following definition:

\begin{Coqsrc}
Definition succ (alpha : t) :=
  match alpha with
    inl n => inl (S n)
  | inr n => inr (S n)
  end.
\end{Coqsrc}

\begin{Coqsrc}
Lemma Successor_succ alpha : Successor (succ alpha) alpha.

Lemma Successor_correct alpha beta : Successor beta alpha <->
                                     beta = succ alpha.
\end{Coqsrc}

We can also check whether an ordinal is a successor:

Finally the nature of any ordinal (inside this notation system) is decidable:

\begin{Coqsrc}
Lemma ZLS_dec (alpha : t) :
  {alpha = 0} +
  {Limit alpha} +
  {beta : t | Successor alpha beta}.
\end{Coqsrc}

@@

It is sometimes useful to compare two ordinal notations with respect to expressive power
(the segment of ordinals  they represent).

The following class specifies a strict inclusion of segments. The notation \texttt{OA} describes a segment $[0,alpha($, and \texttt{OB} is a larger segment (which contains a notation for $\alpha$, whilst $\alpha$ is not represented by any term of type $A$). We require also  that the comparison functions of the two notation systems are compatible.

@@

For instance, we prove that \texttt{Omega} is a sub-notation of
\texttt{Omega\_plus\_Omega} (with $\omega$ as the first ``new'' ordinal, and \texttt{fin} as the injection).

\begin{Coqsrc}
Instance Incl : SubON Omega Omega_plus_Omega omega fin.
\end{Coqsrc}

\subsection{Comparing an ordinal notation with Schütte's model}

Finally, it may be interesting to compare an ordinal notation with the more theoretical model from Schütte (well, at least with our formalization of that model). This would be a relative proof of correctenss of our ordinal  notations.

@@

For instance, the following theorem tells that \texttt{Epsilon0}, our notation system for the segment $[0,\epsilon0)$ is correct with respect to the countqable ordinal $\epsilon_0$
(see chapter~\ref{chap:schutte} for more details).

@@

\index{Projects}

\begin{project}
  When you have read Chapter~\ref{chap:schutte}, prove that the sum of two ordinal notations \texttt{ON\_plus} implements the addition of ordinals.
\end{project}

\section{A notation for finite ordinals}

Let $n$ be some natural number. The segment associated with $n$ is the interval 
$[0,n)\,=\,\{0,1,\dots,n-1\}$.

@@

For more information about this pattern, please look at the numerous mailing lists and 
FAQs on \coq{}).

@@

\section{Isomorphism of ordinal notations}

@@

\begin{exercise}
Let $i$ be some natural number. Prove that the notation systems 
\texttt{Omega} and \texttt{ON\_plus (OrdFin $i$) Omega} are isomorphic.

{\it \textbf{Note:} This property reflects the equality $i+\omega=\omega$ we will prove in larger notation systems, as well as in Schütte's model}
\end{exercise}

%%%% ICI ICI

\begin{itemize}
\item $\omega^n$ (for some  integer  $n\geq 2$) : the type of $n$-uples of natural numbers, with the lexicographic product of $n$ copies of \texttt{Nat.lt}.
\item  $\omega^\omega$: the set of nonincreasing sequences of natural numbers, lexicographically ordered (also the set of finite multisets of natural numbers).
\item $\epsilon_0$: The set of terms in Cantor normal form (see Chap.~\ref{chap:T1}).
\item $\Gamma_0$: The set of terms in Veblen normal form.
\end{itemize}

\begin{todo}
The following section has moved from chapter on hydras.
Adapt the redaction to the $\omega^2$ ordinal segment.
\end{todo}
